To Link against EiC you first need to build the source distribution.
Then linking to EiC from aother programs is done by linking against
the EiC libraries (\T{libeic} and \T{libstdClib}) in \T{EiC/lib}. In
the directory \T{EiC/main/examples} there is an example program called
\T{embedEiC.c} that links to EiC. Build and run it from the
\T{EiC/main/examples} directory by entering (assuming EiC has been
installed in \T{/usr/local/EiC}):

\begin{production}
\begin{verbatim}
% gcc embedEiC.c -L/usr/local/EiC/lib -leic -lstdClib -lm  
% a.out
\end{verbatim}
\end{production}	
 For  communicating commands to EiC from another program there are two
functions supplied:
\begin{production}
\begin{verbatim}
int EiC_run(int argc, char **argv);
\end{verbatim}
\end{production}
 and
\begin{production}
\begin{verbatim}
void EiC_parseString(char *command, ...);
\end{verbatim}
\end{production}

The \T{EiC\_run} function is used to run C source files. The
\T{EiC\_parseString} function is used to pass C or preprocessor commands
to EiC via a string, such as:
\begin{production}
\begin{verbatim}
EiC_parseString("#include <stdio.h>");
EiC_parseString("int a = 10,i;");
EiC_parseString("for(i=0;i<a;i++)"
                "	printf(\"%%d\\n\",i);");
\end{verbatim}
\end{production}
At present the main facility for sharing data between EiC and
other applications is via the address operator \T{@}:

\begin{quote}
\begin{verbatim}
int a @ dddd;
\end{verbatim}
\end{quote}

The above defines \T{a} to be an integer and is stored at address \T{dddd},
which must be an integral constant. The constant address \T{dddd} is
not simply an address conjured up. Its purpose is to enable access to
data, or even functions, defined in compiled code. 

When applied to function definitions, the limitation at this stage is
that the function must take void arguments and return void:

\begin{production}
\begin{verbatim}
     void foo(void) @ dddd; 
\end{verbatim}
\end{production}

The above defines \T{foo} to be a builtin function located at address
\T{dddd}. For example:

\begin{quote}
\begin{verbatim}
int foo[5] = {1,2,3,4,5};
void fooey(void) {printf("fooey called\n");}
     ....
EiC_parseString("int foo[5] @ %ld;", (long)foo);
EiC_parseString("void fooey(void) @ %ld;", (long)fooey);
\end{verbatim}
\end{quote}

Further, \verb+int foo[5] @ 1256;+ defines \T{foo} to be an array of 5
ints mapped at the specified virtual address and the usual pointer
safety rules apply; that is, \verb/foo[5];/ will be caught 
as an illegal operation.


Also, you can pass in data to EiC via setting variables and you can
get EiC to output data to a file. In a future release of EiC, more 
facilities are expected to be added for sharing data between EiC and its
embedding system.

With respect to \T{EiC\_run}, to run the file "myfile.c"  and pass it the
command line arguments "hello" and "world", the following sequence of
commands would be used.

\begin{production}
\begin{verbatim}
char *argv[] = {"myfile.c", "hello", "world"};
int  argc = sizeof(argv)/sizeof(char*);
EiC_run(argc, argv);
\end{verbatim}
\end{production}
